---
title: "Seurat CPU Harmonized Benchmark"
author: "Edgar (Seyun) Jeong"
date: "`r Sys.Date()`"
output: html_document
---

# Seurat CPU Harmonized Pipeline

This notebook is intentionally written to minimize *avoidable* divergence from
the Scanpy / rapids-singlecell reference while staying inside Seurat's API.

**Key differences that remain even here:**

1. Seurat still requires scaled data before RunPCA on selected features.
2. Seurat's Leiden wrapper exposes `n.start` / `n.iter` semantics that are not
   the same as Scanpy's or cuGraph's iteration parameters.
3. Even when clustering on the NN graph instead of the usual SNN graph,
   you are still using Seurat's graph object and Seurat's wrapper.

**Key harmonization moves implemented here:**

- No hidden filtering during object construction
- Same cells and genes as the canonical allowlists
- Exact CPU nearest-neighbor search via RANN (`nn.eps = 0`)
- Clustering on the NN graph (not the default SNN graph)
- Reduced DE prefiltering: `min.pct = 0`, `logfc.threshold = 0`

## Setup
```{r}
Sys.setenv(RETICULATE_PYTHON = "/home/seyun/anaconda3/bin/python")
library(reticulate)
use_python("/home/seyun/anaconda3/bin/python", required = TRUE)
```


```{r setup, message=FALSE, warning=FALSE}
# === CHANGE THIS TO SWITCH DATASETS ===
dataset_key <- "lung65k"   # or "lung65k"
config_path <- "benchmark_config.json"
```

```{r libraries, message=FALSE, warning=FALSE}
library(Seurat)
library(Matrix)
library(jsonlite)

cat("Seurat version:", as.character(packageVersion("Seurat")), "\n")
cat("SeuratObject version:", as.character(packageVersion("SeuratObject")), "\n")
cat("Matrix version:", as.character(packageVersion("Matrix")), "\n")
cat("R version:", paste(R.version$major, R.version$minor, sep = "."), "\n")
```

## Load Config

```{r config}
`%||%` <- function(x, y) if (is.null(x)) y else x

cfg <- fromJSON(config_path, simplifyVector = TRUE)
stopifnot(dataset_key %in% names(cfg$datasets))

gcfg <- cfg$global
dcfg <- cfg$datasets[[dataset_key]]

out_prefix <- file.path("write", paste0(dcfg$pipeline_prefix, "_seurat_cpu_harmonized"))
dir.create("write", showWarnings = FALSE, recursive = TRUE)

out_paths <- list(
  rds              = paste0(out_prefix, ".rds"),
  clusters         = paste0(out_prefix, "_clusters.csv"),
  markers          = paste0(out_prefix, "_markers.csv"),
  markers_filtered = paste0(out_prefix, "_markers_filtered.csv"),
  umap             = paste0(out_prefix, "_umap.csv"),
  hvg              = paste0(out_prefix, "_hvg.csv"),
  spec             = paste0(out_prefix, "_spec.json")
)

cat("Dataset          :", dcfg$name, "\n")
cat("Raw input        :", dcfg$input_path, "\n")
cat("Canonical cells  :", dcfg$canonical_cells_txt, "\n")
cat("Canonical genes  :", dcfg$canonical_genes_txt, "\n")
```

## Helper Functions

```{r helpers}
read_allowlist <- function(path) {
  x <- readLines(path, warn = FALSE)
  x[nzchar(x)]
}

load_h5ad_counts <- function(h5ad_path) {
  library(reticulate)
  anndata <- import("anndata")
  
  ad <- anndata$read_h5ad(h5ad_path)
  
  layer_keys <- tryCatch(
    as.character(py_to_r(ad$layers$keys())),
    error = function(e) character()
  )
  if ("counts" %in% layer_keys) {
    x <- ad$layers$get("counts")
  } else {
    x <- ad$X
  }
  
  # h5ad: cells x genes → Seurat: genes x cells
  x <- x$transpose()$tocsc()
  
  p    <- as.integer(as.vector(py_to_r(x$indptr)))
  i    <- as.integer(as.vector(py_to_r(x$indices)))
  vals <- as.numeric(as.vector(py_to_r(x$data)))
  dims <- as.integer(as.vector(py_to_r(x$shape)))
  
  gene_names <- as.character(as.vector(py_to_r(ad$var_names$to_list())))
  cell_names <- as.character(as.vector(py_to_r(ad$obs_names$to_list())))
  
  counts <- new("dgCMatrix", p = p, i = i, x = vals, Dim = dims)
  rownames(counts) <- make.unique(gene_names)
  colnames(counts) <- cell_names
  counts
}

load_counts <- function(ds) {
  if (ds$input_type == "10x_mtx") {
    counts <- Read10X(data.dir = ds$input_path)
    rownames(counts) <- make.unique(rownames(counts))
    return(counts)
  }
  if (ds$input_type == "h5ad") {
    return(load_h5ad_counts(ds$input_path))
  }
  stop("Unsupported input_type: ", ds$input_type)
}

subset_to_canonical <- function(counts, allow_genes, allow_cells) {
  common_genes <- intersect(allow_genes, rownames(counts))
  common_cells <- intersect(allow_cells, colnames(counts))

  if (length(common_genes) == 0) stop("No overlapping genes between raw input and canonical gene allowlist.")
  if (length(common_cells) == 0) stop("No overlapping cells between raw input and canonical cell allowlist.")

  counts[common_genes, common_cells, drop = FALSE]
}
```

## Load and Subset to Canonical Input

```{r load_data}
allow_cells <- read_allowlist(dcfg$canonical_cells_txt)
allow_genes <- read_allowlist(dcfg$canonical_genes_txt)
counts <- Read10X(data.dir = "write/lung_65k_canonical_10x/")
counts <- subset_to_canonical(counts, allow_genes, allow_cells)

cat("Counts after canonical subset:", ncol(counts), "cells ×", nrow(counts), "genes\n")
```

```{r create_object}
obj <- CreateSeuratObject(
  counts       = counts,
  project      = dcfg$name,
  min.cells    = 0,    # No hidden filtering — canonical input is already filtered
  min.features = 0
)

obj[["percent.mt"]]   <- PercentageFeatureSet(obj, pattern = "^MT-")
obj[["percent.ribo"]] <- PercentageFeatureSet(obj, pattern = "^RP[SL]")
obj[["percent.hb"]]   <- PercentageFeatureSet(obj, pattern = "^HB[^(P)]")

cat("Seurat object:", ncol(obj), "cells ×", nrow(obj), "genes\n")
```

## Normalize

```{r normalize}
system.time({
  obj <- NormalizeData(
    obj,
    normalization.method = "LogNormalize",
    scale.factor         = gcfg$target_sum,
    verbose              = FALSE
  )
})
```

## HVG Selection

```{r hvg}
system.time({
  obj <- FindVariableFeatures(
    obj,
    selection.method = "vst",
    nfeatures        = dcfg$n_top_genes,
    verbose          = FALSE
  )
})
cat("HVGs selected:", length(VariableFeatures(obj)), "\n")
```

## Scale Data

**Seurat requires this before PCA. Scanpy and rsc benchmarks skip scaling.**
This is an irreducible methodological difference.

```{r scale}
system.time({
  obj <- ScaleData(
    obj,
    features = VariableFeatures(obj),
    verbose  = FALSE
  )
})
```

## PCA

```{r pca}
system.time({
  obj <- RunPCA(
    obj,
    features = VariableFeatures(obj),
    npcs     = gcfg$pca_n_comps,
    seed.use = gcfg$random_state,
    verbose  = FALSE
  )
})
```

## Neighbors

Uses exact RANN search (`nn.eps = 0`) instead of Annoy.
If `graph_mode == "nn"`, clusters on the NN graph directly to reduce
divergence from Scanpy / rapids-singlecell KNN-graph clustering.

```{r neighbors}
graph_mode <- gcfg$seurat_graph_mode

system.time({
  if (isTRUE(graph_mode == "nn")) {
    obj <- FindNeighbors(
      obj,
      reduction      = "pca",
      dims           = seq_len(dcfg$neighbors_n_pcs),
      k.param        = dcfg$n_neighbors,
      return.neighbor = FALSE,
      compute.SNN    = FALSE,
      nn.method      = gcfg$seurat_nn_method,
      nn.eps         = gcfg$seurat_nn_eps,
      graph.name     = "RNA_nn",
      verbose        = FALSE
    )
    graph_for_clustering <- "RNA_nn"
  } else {
    obj <- FindNeighbors(
      obj,
      reduction      = "pca",
      dims           = seq_len(dcfg$neighbors_n_pcs),
      k.param        = dcfg$n_neighbors,
      return.neighbor = FALSE,
      compute.SNN    = TRUE,
      prune.SNN      = gcfg$seurat_prune_snn,
      nn.method      = gcfg$seurat_nn_method,
      nn.eps         = gcfg$seurat_nn_eps,
      graph.name     = c("RNA_nn", "RNA_snn"),
      verbose        = FALSE
    )
    graph_for_clustering <- "RNA_snn"
  }
})

cat("Graph mode:", graph_mode, "\n")
cat("Clustering graph:", graph_for_clustering, "\n")
```

## Leiden Clustering

Seurat semantics differ from Scanpy/cuGraph.
`n.start = 1` is pinned to reduce an extra search dimension that
Scanpy/rsc do not expose in the same way.

```{r leiden}
system.time({
  obj <- FindClusters(
    obj,
    graph.name    = graph_for_clustering,
    resolution    = dcfg$leiden_resolution,
    algorithm     = 4,
    leiden_method  = gcfg$seurat_leiden_method,
    n.start       = gcfg$seurat_leiden_n_start,
    n.iter        = gcfg$seurat_leiden_n_iter,
    random.seed   = gcfg$random_state,
    verbose       = FALSE
  )
})

cat("Leiden clusters:", length(levels(Idents(obj))), "\n")
```

## UMAP

For visualization only — not the primary validation target.

```{r umap}
system.time({
  obj <- RunUMAP(
    obj,
    reduction   = "pca",
    dims        = seq_len(dcfg$neighbors_n_pcs),
    umap.method = "uwot",
    n.neighbors = dcfg$n_neighbors,
    metric      = gcfg$neighbor_metric,
    min.dist    = gcfg$umap_min_dist,
    spread      = gcfg$umap_spread,
    seed.use    = gcfg$random_state,
    verbose     = FALSE
  )
})
```

```{r umap_plot, fig.width=8, fig.height=6}
DimPlot(obj, reduction = "umap", label = TRUE, pt.size = 0.5) +
  ggtitle(paste0(dcfg$name, " — Seurat CPU Harmonized"))
```

## Differential Expression

Minimize Seurat-side DE prefiltering (`min.pct = 0`, `logfc.threshold = 0`)
so that downstream comparison applies the same `p_adj`/`logFC` threshold
externally across all libraries.

```{r de, warning=FALSE}
presto_installed <- requireNamespace("presto", quietly = TRUE)
cat("presto installed:", presto_installed, "\n")

system.time({
  markers <- FindAllMarkers(
    obj,
    only.pos        = FALSE,
    test.use        = "wilcox",
    min.pct         = 0,
    logfc.threshold = 0,
    verbose         = FALSE
  )
})

markers_filtered <- subset(markers, p_val_adj < 0.05 & avg_log2FC > 0.1)

cat("Total marker rows:", nrow(markers), "\n")
cat("Filtered markers (p_val_adj < 0.05 & avg_log2FC > 0.1):", nrow(markers_filtered), "\n")
```

## Save All Outputs

```{r save}
# RDS
saveRDS(obj, out_paths$rds)

# Cluster assignments
write.csv(
  data.frame(
    barcode = colnames(obj),
    leiden  = as.character(Idents(obj)),
    stringsAsFactors = FALSE
  ),
  out_paths$clusters,
  row.names = FALSE
)

# DE markers
write.csv(markers, out_paths$markers, row.names = FALSE)
write.csv(markers_filtered, out_paths$markers_filtered, row.names = FALSE)

# UMAP coordinates
write.csv(
  data.frame(
    barcode = rownames(Embeddings(obj, "umap")),
    Embeddings(obj, "umap"),
    stringsAsFactors = FALSE
  ),
  out_paths$umap,
  row.names = FALSE
)

# HVG list
write.csv(
  data.frame(gene = VariableFeatures(obj), stringsAsFactors = FALSE),
  out_paths$hvg,
  row.names = FALSE
)

cat("Saved all CSV outputs.\n")
```

## Spec JSON

```{r spec}
spec <- list(
  pipeline              = "seurat_cpu_harmonized",
  dataset               = dcfg$name,
  seurat_version        = as.character(packageVersion("Seurat")),
  seurat_object_version = as.character(packageVersion("SeuratObject")),
  matrix_version        = as.character(packageVersion("Matrix")),
  input                 = dcfg$input_path,
  canonical_cells       = dcfg$canonical_cells_txt,
  canonical_genes       = dcfg$canonical_genes_txt,
  parameters = list(
    target_sum         = gcfg$target_sum,
    hvg_method         = "vst",
    n_top_genes        = dcfg$n_top_genes,
    scale_data         = TRUE,
    scale_features     = "VariableFeatures(obj)",
    pca_n_comps        = gcfg$pca_n_comps,
    n_neighbors        = dcfg$n_neighbors,
    neighbors_n_pcs    = dcfg$neighbors_n_pcs,
    neighbor_metric    = gcfg$neighbor_metric,
    nn_method          = gcfg$seurat_nn_method,
    nn_eps             = gcfg$seurat_nn_eps,
    graph_mode         = graph_mode,
    clustering_graph   = graph_for_clustering,
    clustering_algorithm = "Leiden",
    leiden_method      = gcfg$seurat_leiden_method,
    leiden_n_start     = gcfg$seurat_leiden_n_start,
    leiden_n_iter      = gcfg$seurat_leiden_n_iter,
    leiden_resolution  = dcfg$leiden_resolution,
    umap_method        = "uwot",
    umap_n_neighbors   = dcfg$n_neighbors,
    umap_metric        = gcfg$neighbor_metric,
    umap_min_dist      = gcfg$umap_min_dist,
    umap_spread        = gcfg$umap_spread,
    random_state       = gcfg$random_state,
    de_test            = "wilcox",
    de_min_pct         = 0,
    de_logfc_threshold = 0,
    presto_installed   = presto_installed
  ),
  results = list(
    n_cells              = ncol(obj),
    n_genes              = nrow(obj),
    n_hvg                = length(VariableFeatures(obj)),
    n_clusters           = length(levels(Idents(obj))),
    n_marker_rows        = nrow(markers),
    n_marker_rows_filtered = nrow(markers_filtered)
  )
)

writeLines(toJSON(spec, pretty = TRUE, auto_unbox = TRUE), out_paths$spec)
cat("Spec written to:", out_paths$spec, "\n")
```

## Summary

```{r summary}
cat(strrep("=", 60), "\n")
cat("RESULTS SUMMARY\n")
cat(strrep("=", 60), "\n")
cat("Dataset         :", dcfg$name, "\n")
cat("Cells           :", ncol(obj), "\n")
cat("Genes           :", nrow(obj), "\n")
cat("HVGs            :", length(VariableFeatures(obj)), "\n")
cat("Clusters        :", length(levels(Idents(obj))), "\n")
cat("Marker rows     :", nrow(markers), "\n")
cat("Filtered markers:", nrow(markers_filtered), "\n")
cat("\nOutput files:\n")
for (p in out_paths) cat("  ", p, "\n")
```

```{r session_info}
sessionInfo()
```
